Interpreting Random Forest Predictions for Bullet Matching Using LIME (?)

### 2018-10-15 paper/paper.Rmd
updated code in paper to save important objects (again)

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

# Load libraries
library(dplyr)
library(ggplot2)
library(lime)
```

@@

Bullet matching is a practice that has historically been performed by hand by specially trained firearm examiners. The examiners visually assess the striation markings on the bullets under a high powered microscope to determine whether the bullets were fired from the same gun barrel. In more recent times, the scientific community has been calling for the inclusion of more data driven techniques to be used in forensic investigations that would allow for the reporting of an uncertainty along with the conclusion drawn from the analysis.

@@

@@

@@

@@

```{r}
# Read in the Hamby 224 combined test data and explanations
hamby224_test_explain <- readRDS("../data/hamby224_test_explain.rds")
```

```{r}
hamby224_test_explain %>%
  filter(set == "1") %>%
  select(case, bullet1:land2, rfscore) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2)) +
  geom_tile(aes(fill = rfscore)) +
  facet_grid(bullet1 ~ bullet2) +
  theme_bw() +
  scale_fill_gradient2(low = "grey", high = "orange", midpoint = 0.5)

hamby224_test_explain %>%
  filter(set == "1") %>%
  select(case, bullet1:land2, mypred) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2)) +
  geom_tile(aes(fill = mypred)) +
  facet_grid(bullet1 ~ bullet2) +
  theme_bw() +
  scale_fill_gradient2(low = "grey", high = "orange", midpoint = 0.5)

hamby224_test_explain %>%
  filter(set == "1") %>%
  select(case, bullet1:land2, label_prob) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2)) +
  geom_tile(aes(fill = label_prob)) +
  facet_grid(bullet1 ~ bullet2) +
  theme_bw() +
  scale_fill_gradient2(low = "grey", high = "orange", midpoint = 0.5)

hamby224_test_explain %>%
  filter(set == "1") %>%
  select(case, bullet1:land2, samesource) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2)) +
  geom_tile(aes(fill = samesource)) +
  facet_grid(bullet1 ~ bullet2) +
  theme_bw()
```

```{r, fig.width = 8, fig.height = 4}
hamby224_explain <- readRDS("../data/hamby224_explain.rds")

lime::plot_explanations(hamby224_explain)

hamby224_test_explain %>%
  filter(set == "1") %>%
  ggplot(aes(x = case, y = feature_desc)) + 
  geom_tile(aes(fill = feature_weight)) + 
  facet_grid(feature ~ label, space = "free", scale = "free") + 
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2(midpoint = 0)

hamby224_test_explain %>%
  filter(set == "1", samesource == "FALSE") %>%
  ggplot(aes(x = case, y = feature_desc)) + 
  geom_tile(aes(fill = feature_weight)) + 
  facet_grid(feature ~ bullet1, space = "free", scale = "free") + 
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2(midpoint = 0)

hamby224_test_explain %>%
  filter(set == "1", samesource == "FALSE") %>%
  ggplot(aes(x = case, y = feature_desc)) + 
  geom_tile(aes(fill = feature_weight)) + 
  facet_grid(bullet1 ~ bullet2, space = "free", scale = "free") + 
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2(midpoint = 0) +
  theme(legend.position = "bottom")
```

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, cache = TRUE)

# Load libraries
library(tidyverse)
library(bulletr)
library(lime)
library(randomForest)
```

@@

The process of matching bullets to the gun that shot the bullet has traditionally been performed by hand by specially trained firearm examiners. The examiners visually assess the striation markings on the bullets under a high powered microscope to determine whether the bullets were fired from the same gun barrel. Recently, the scientific community has been encouraging the inclusion of more data driven techniques to be used in forensic investigations that would allow for the reporting of an uncertainty along with the conclusion drawn from the analysis.

@@

```{r}
# Load in the training data (Hamby Data 173 and 252)
hamby173and252_train <- read.csv("../data/hamby173and252_train.csv")
```

@@

```{r}
# Load in the Hamby 224 datasets
hamby224_set1 <- readRDS("../data/h224-set1-features.rds")
hamby224_set11 <- readRDS("../data/h224-set11-features.rds")
```

```{r}
# Obtain features used when fitting the rtrees random forest
rf_features <- rownames(rtrees$importance)

# Clean the Hamby 224 set 1 data
hamby224_set1_cleaned <- hamby224_set1 %>%
  select(-bullet_score, -land1, -land2, -aligned, -striae, -features) %>%
  rename(bullet1 = bulletA,
         bullet2 = bulletB, 
         land1 = landA,
         land2 = landB) %>%
  mutate(study = factor("Hamby_224"), 
         set = factor("1"),
         bullet1 = factor(bullet1),
         bullet2 = factor(bullet2),
         land1 = factor(land1),
         land2 = factor(land2)) %>%
  select(study, set, bullet1:land2, rf_features, rfscore, samesource)

# Clean the Hamby 224 set 11 data
hamby224_set11_cleaned <- hamby224_set11 %>%
  select(-bullet_score, -land1, -land2, -aligned, -striae, -features) %>%
  rename(bullet1 = bulletA,
         bullet2 = bulletB, 
         land1 = landA,
         land2 = landB) %>%
  mutate(study = factor("Hamby_224"), 
         set = factor("2"),
         bullet1 = recode(factor(bullet1), 
                          "Bullet 1" = "1", 
                          "Bullet 2" = "2", 
                          "Bullet I" = "I"),
         bullet2 = recode(factor(bullet2), 
                          "Bullet 1" = "1", 
                          "Bullet 2" = "2", 
                          "Bullet I" = "I"),
         land1 = recode(factor(land1), 
                        "Land 1" = "1", "Land 2" = "2", "Land 3" = "3", 
                        "Land 4" = "4", "Land 5" = "5", "Land 6" = "6"),
         land2 = recode(factor(land2), 
                        "Land 1" = "1", "Land 2" = "2", "Land 3" = "3", 
                        "Land 4" = "4", "Land 5" = "5", "Land 6" = "6")) %>%
  select(study, set, bullet1:land2, rf_features, rfscore, samesource)

# Join the two cleaned Hamby 224 sets into one testing set
hamby224_test <- suppressWarnings(bind_rows(hamby224_set1_cleaned,
                                            hamby224_set11_cleaned))

# Save the test data as a .csv file
write.csv(hamby224_test, "../data/hamby224_test.csv", row.names = FALSE)
```

@@

```{r echo = FALSE}
# Apply the lime function from the lime package (with a seed set)
# Note that the as_classifier must be added since rtrees is from the
# randomForest package and not fit using caret or one of the other 
# available models specified in the lime package. Additionally, the
# randomForest package must be loaded in order to run the functions
# from the lime package. (I should check on exactly how this works.)
set.seed(84902)
hamby224_lime <- lime(x = hamby173and252_train %>% select(rf_features),
                      model = as_classifier(rtrees))

# Save the lime object
saveRDS(hamby224_lime, "../data/hamby224_lime.rds")
```

```{r}
# Apply the explain function from the lime package (with a seed set)
set.seed(84902)
hamby224_explain <- lime::explain(hamby224_test %>% select(rf_features), 
                                  hamby224_lime, 
                                  n_labels = 1, 
                                  n_features = 3)

# Save the explainer object
saveRDS(hamby224_explain, "../data/hamby224_explain.rds")
```

```{r echo = FALSE}
# Add a case variable to the test data
hamby224_test <- hamby224_test %>%
  mutate(case = as.character(1:dim(hamby224_test)[1])) %>%
  select(case, study:samesource)

# Obtain features used when fitting the rtrees random forest
rf_features <- rownames(rtrees$importance)

# Join the data and the explanations and edit and add additional variables
hamby224_test_explain <- full_join(hamby224_test, hamby224_explain, by = "case") %>%
  mutate(case = factor(case),
         set = factor(set),
         bullet1 = factor(bullet1),
         bullet2 = factor(bullet2),
         land1 = factor(land1),
         land2 = factor(land2)) %>%
  mutate(mypred = rep(predict(rtrees, hamby224_test %>% select(rf_features), 
                              type = "prob")[,2], each = 3)) %>%
  select(case:rfscore, mypred, samesource:prediction) %>%
  mutate(feature_desc = factor(feature_desc),
         feature = factor(feature)) %>%
  mutate(feature_number = readr::parse_number(feature_desc),
         strictly_less = FALSE)

# Finish creating the strictly less than variable
hamby224_test_explain$strictly_less[grep("< ", hamby224_test_explain$feature_desc)] <- TRUE

# Reorder the variable of feature_desc for plotting purposes
hamby224_test_explain <- hamby224_test_explain %>%
  mutate(
    feature_desc = reorder(feature_desc, strictly_less),
    feature_desc = reorder(feature_desc, feature_number),
    feature_desc = reorder(feature_desc, as.numeric(feature))
  )

# Save the combined test and explain data
saveRDS(hamby224_test_explain, "../data/hamby224_test_explain.rds")
```

@@

```{r}
hamby224_test_explain %>%
  ggplot(aes(x = feature_desc)) + 
  geom_bar() +
  coord_flip()
```

```{r}
hamby224_test_explain %>%
  filter(set == "1") %>%
  select(case, bullet1:land2, rfscore) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2)) +
  geom_tile(aes(fill = rfscore)) +
  facet_grid(bullet1 ~ bullet2) +
  theme_bw() +
  scale_fill_gradient2(low = "grey", high = "orange", midpoint = 0.5)
```

```{r}
hamby224_test_explain %>%
  filter(set == "1") %>%
  select(case, bullet1:land2, samesource) %>%
  distinct() %>%
  ggplot(aes(x = land1, y = land2)) +
  geom_tile(aes(fill = samesource)) +
  facet_grid(bullet1 ~ bullet2) +
  theme_bw()
```

```{r, fig.width = 8, fig.height = 4}
lime::plot_explanations(hamby224_explain)
```

```{r}
hamby224_test_explain %>%
  filter(set == "1") %>%
  ggplot(aes(x = case, y = feature_desc)) + 
  geom_tile(aes(fill = feature_weight)) + 
  facet_grid(feature ~ label, space = "free", scale = "free") + 
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2(midpoint = 0) +
  theme(legend.position = "bottom") +
  ylab("Feature Sets") 
```

```{r}
hamby224_test_explain %>%
  filter(set == "1", samesource == "FALSE") %>%
  ggplot(aes(x = case, y = feature_desc)) + 
  geom_tile(aes(fill = feature_weight)) + 
  facet_grid(feature ~ bullet1, space = "free", scale = "free") + 
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2(midpoint = 0)
```

```{r}
hamby224_test_explain %>%
  filter(set == "1", samesource == "FALSE") %>%
  ggplot(aes(x = case, y = feature_desc)) + 
  geom_tile(aes(fill = feature_weight)) + 
  facet_grid(bullet1 ~ bullet2, space = "free", scale = "free") + 
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_gradient2(midpoint = 0) +
  theme(legend.position = "bottom")
```
